\toolname: A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-04-25 doc/paper.tex
Changes to entire paper. Too many issues fixed to list here.

\usepackage{textcomp} \newcommand{\toolname}{Perquimans?!} \begin{document}
	\title{A Tool for Visualizing Patterns of Spreadsheet Function Combinations}

@@

	% As a general rule, do not put math, special symbols or citations
	% in the abstract
	\begin{abstract} Spreadsheet environments often come equipped with an abundance
		of functions and operations to manipulate data, but it can be difficult to
		understand how programmers actually use these in practice. Furthermore, users
		can combine several functions into a complex formula, complicating matters for
		both researchers and practitioners who want to study formulae to improve
		spreadsheet practices. Therefore, we developed a tool that visualizes patterns
		of function combination in spreadsheets as an interactive tree of
		possibilities. Using spreadsheets from the public spreadsheet corpora, we then
		apply to the tool to real datasets and demonstrate its ability to capture the
		most common and most anomalous patterns of function combination and their
		contexts in actual workbooks. \end{abstract}

@@

	Alongside the tool, we discuss several of the key decisions we faced during
	design, such as what to do with redundant formulae in a sheet or how to
	represent optional arguments, and we describe how our overall design goals both
	helped and hindered this process. We then apply the tool to a number of case
	studies to evaluate how much it could contribute to several different contexts:

	\begin{itemize} \item Detection of code smells \item
		Spreadsheet education \end{itemize}

@@

	\section{Case Study} The obvious question to pose to any visualization is this:
	What do the images tell us that the text could not? It is not enough to prove
	that something can be visualized; we must also demonstrate what we can be
	gained from any particular visualization. As such, we raised a few tasks in the
	introduction in which we thought the tool could help: identifying places for
	API improvement, detecting bad smells, and guiding spreadsheet education. To
	demonstrate the tool's applicability, we sought out places in the tree which
	best exemplify these tasks.

@@

	%Lookup example
	Otherwise, the LOOKUP functions are particularly notorious for their confusing
	mechanisms and, as such, have warranted some especial attention with regards to
	spreadsheet design. In a recent study, Hermans, Aivaloglou, and Jansen probed
	the Enron dataset for uses of VLOOKUP and HLOOKUP specifically. The issue at
	hand was the optional fourth argument: when omitted or set to TRUE, the
	function attempts to return a value from another column by matching two values
	approximately, but when set to FALSE, the match must be exact. Furthermore,
	exact matching works only with sorted columns or rows; if the programmer
	neglects to sort values before calling the *LOOKUP function, they may acquire
	inaccurate results \cite{hermans2015detecting}. \par

	Given access to the same dataset, how might this tool approach the issue? On
	the surface, the tool supports easy quantification through the tooltips
	displayed for each node, as shown in figure \ref{fig:vhlookups}, answering the
	question, "How often do spreadsheet programmers use this function?". But, more
	to the point, the tool's distinction between the numbers of arguments for a
	single function lets the user compare frequencies of three-argument forms with
	the four-argument. The visualization does not, however, discern between the
	different values for booleans in the tree, much less whether the spreadsheet
	columns are ordered -- but the linking of examples to patterns allows quick
	insight into which files are at risk for the problems they outlined. \par

	\begin{figure}[h] \centering \includegraphics[width=.5\textwidth]{vlookup}
		\includegraphics[width=.225\textwidth]{vlookupbox}
		\includegraphics[width=.225\textwidth]{hlookupbox}
		\includegraphics[width=.5\textwidth]{hlookup} \label{fig:vhlookups}
		\caption{VLOOKUP and HLOOKUP with four arguments. Numbers not expected to
			correspond perfectly with \cite{hermans2015detecting}.} \end{figure}

	Also at hand are issues with the standard LOOKUP function. Unlike the two
	variations above, LOOKUP accepts either two or three arguments, depending on
	which API-defined form the programmer uses: the vector form or the array form,
	the primary difference being whether the second parameter is a one-column/-row
	range or a two-dimensional array. The array form, as it happens, is strongly
	discouraged by the API in favor of the V/HLOOKUP, a notion that might also be
	leveraged in the pursuit in good spreadsheet design.

	\begin{figure}[h] \centering \includegraphics[width=.5\textwidth]{lookup-2}
		\includegraphics[width=.5\textwidth]{lookup-3} \label{fig:lookups}
		\caption{LOOKUP with two (6 occurrences) and three (1648) arguments.}
	\end{figure}

@@

	\section{User Study} During development, we conducted a brief, four-participant
	user study in order to detect conflicts between our design and the goals in
	section \ref{goals}. We designed the study to be exploratory; given trees based
	in the Enron dataset and at least twenty minutes, we asked the participants to
	adopt the persona of a consultant evaluating a company's spreadsheet practice,
	allowing for questions like which functions on which the employees most
	depended or which sheets or employees presented the most anomalous or dangerous
	designs. Though we instructed them on how to use the tool and suggested that
	they visit one of the most populous trees in the set, we put aside any specific
	tasks or questions and let them choose their own paths. Though interpretation
	of the phrase ``spreadsheet practice" can certainly vary from participant to
	participant, we readily accepted this interpretive flexibility since we wanted
	a range of views and backgrounds despite the persona. We clarified,
	furthermore, that their observations could be directed either at the quality of
	the data conveyed by the tool or at the tool's design itself. \par

@@

	\subsection{Informative Qualification}

	\section{Future Work} For these user and case studies, we relied solely on
	Enron's spreadsheets to find our results. However, Enron's focus on finances
	and energy represents only one context in which people use spreadsheets. Other
	corpora, like EUSES and Fuse, draw from other domains and industries, and, as
	Jansen found in his comparison of spreadsheets Enron and EUSES, different
	companies rely on different functions. Though the use of a single corpora
	nevertheless suffices for evaluation, we may yet be able to find more
	interesting applications and uses for the tool in other types of spreadsheet.
	\par

	Additionally, Excel is only one of many different spreadsheet products
	available, chosen for this project because of its prevalence. But other
	spreadsheets, such as Gnumeric\footnote{http://www.gnumeric.org/} or Google
	Sheets \footnote{https://www.google.com/sheets/about/} represent other
	approaches to spreadsheet systems. Even within identical domains, there's no
	guarantee that practices and functions will hold across these varied offerings.
	This, then, warrants further analysis, in which good visualization might serve
	well. \par

	A larger and related problem is the loss of argument combination. For example,
	figure \ref{fig:index} shows the situation of the INDEX tree, wherein each
	argument position has a few distinct possibilities. While it is easy to break
	down, for each position, which argument types are most common (ranges, MATCH,
	and numbers, respectively), the tree is silent on whether MATCH functions are
	used together in the second and third arguments, among other concerns. Though
	these argument combinations could be pieced together by manually inspecting the
	lists of examples, it is nevertheless inconvenient, though not outside the
	realm of future improvement. \par

	\begin{wrapfigure}{O}{.25\textwidth} \centering
		\includegraphics[width=.25\textwidth]{longIF} \label{fig:longif} \caption{The
			many possibilities of IF, argument 3, demonstrate the problem of having too
			many options.} \end{wrapfigure}

	Further contextual information.

@@

	Redesign.

@@

	To be sure, we still have a lot of room for improvement. Though the exploratory
	philosophy fostered certain goals well, such as the ability to discover new
	problematic smells without knowing beforehand what they are, it also hindered
	the project in others. Some users, for example, found the trees of popular
	functions to be too crowded to explore well, or that their intention to use the
	tool to examine the good or the bad design exclusively was not fully supported
	by a tool that claimed to know neither. Many of the problems, however, were not
	essential to this design conflict and will be overcome in future iterations of
	design by refining icons and improving maneuverability. \par

	Either way, the visualization of these large datasets represents an important
	step in improving practice overall. By taking these voluminous oceans of data
	offered in spreadsheet corpora and rendering them comprehensible through
	images, we seek to draw attention to valuable information that would otherwise
	go unnoticed.

>>>>

\usepackage{textcomp} \newcommand{\toolname}{Perquimans?!} \begin{document}
	\title{\toolname: A Tool for Visualizing Patterns of Spreadsheet Function Combinations}

@@

	% As a general rule, do not put math, special symbols or citations
	% in the abstract
	\begin{abstract} Spreadsheet environments often come equipped with an abundance
		of functions and operations to manipulate data, but it can be difficult to
		understand how programmers actually use these functions in practice.
		Furthermore, users can combine several functions into a complex formula,
		complicating matters for both researchers and practitioners who want to study
		formulae to improve spreadsheet practices. Therefore, we developed \toolname, a
		tool that visualizes patterns of function combination in spreadsheets as an
		interactive tree of possibilities, capable of capturing both the most common
		and most anomalous patterns of function combination and their contexts in
		actual workbooks. Using spreadsheets from the Enron corpus, we conduct both a
		case study and several user studies to explore \toolname's various
		applications, particularly those in flexible smell detection and spreadsheet
		education. \end{abstract}

@@

	Alongside the tool, we discuss several of the key decisions we faced during
	design, such as what to do with redundant formulae in a sheet or how to
	represent optional arguments, and we describe how our overall design goals both
	helped and hindered this process. We then apply the tool to a number of case
	studies to evaluate how much it could contribute to several different contexts,
	most notably code smell detection and spreadsheet education.

@@

	\section{Case Study} The obvious question to pose to any visualization is this:
	What do the images tell us that the text could not? It is not enough to prove
	that something can be visualized; we must also demonstrate what we can be
	gained from any particular visualization. As such, we raised a few tasks in the
	introduction in which we thought the tool could help: detecting bad smells and guiding spreadsheet education. To
	demonstrate the tool's applicability, we sought out places in the tree which
	best exemplify these tasks.

@@

@@

	%Lookup example
	For a deeper example, the LOOKUP functions are particularly notorious for their
	confusing mechanisms and, as such, have warranted some extra attention with
	regards to spreadsheet design. In short, these functions are for "when you need
	to look in a single row or column and find a value from the same position in a
	second row or
	column"\footnote{https://support.office.com/en-us/article/LOOKUP-function-446d94af-663b-451d-8251-369d5e3864cb}. Examples of each can be found in Figure \ref{fig:lookupexample}.

	\begin{figure} \centering \includegraphics[width=.5\textwidth]{lookupexample}
		\label{fig:lookupexample} \caption{Demonstrations of LOOKUP and VLOOKUP}
	\end{figure}

	LOOKUP has two forms: vector and array. In either case, the first argument is
	the value which you are trying to match in another column. If the formula is in
	the vector form, that column is the second argument, so in Figure
	\ref{fig:lookupexample}, the formula in cell D2 (\textit{=LOOKUP(3, A2:A6,
		B2:B6)}) will search for the value of 3 between cells A2 and A6. The value the
	formula returns then comes from the corresponding cell in the optional third
	argument's column; since the lookup values matches the value in cell A4, it
	returns the value from cell B4. The array form of the LOOKUP works the same
	way, except that the second argument is a two-dimensional table where it
	matches values in the table's first column and returns values from the
	last\footnote{ Depending on the dimensions of the lookup table, the function
		may substitute rows for columns.}.

	VLOOKUP works closely with this array form, except the user defines in its
	third argument which column from which to return values. In \textit{=VLOOKUP(3,
		A2:B6, 2)}, the argument ``2" tells to return the value from the second column,
	which is \textit{B2:B6} here. Furthermore, it also accepts a boolean value for
	an optional argument: when set to FALSE, it returns only exact matches, and
	when TRUE or left out, it approximates the match to the closest value beneath
	it. We demonstrate this in Figure \ref{fig:lookupexample} in the bottom two
	VLOOKUP functions, one returning ``Friday" and the other an error. HLOOKUP
	works identically but focuses on rows instead of columns.

	All of these variations may seem daunting to a beginner, but \toolname can
	help. Because of its design, users can distinguish between forms of the same
	function, as seen in Figure \ref{fig:lookups} where the one can explore both
	the frequency with which the spreadsheet programmers used these different forms
	-- three-argument vector form much more than the array form -- and then
	retrieve examples of every configuration in the dataset. The same goes for the
	V/HLOOKUP functions, as shown partially in Figure \ref{fig:vhlookups}

	\begin{figure}[h] \centering \includegraphics[width=.5\textwidth]{lookup-2}
		\includegraphics[width=.5\textwidth]{lookup-3} \label{fig:lookups}
		\caption{LOOKUP with two and three arguments.} \end{figure}

	Furthermore, these examples are good as well for exploring how it might bridge
	issues of bad design with education. In a recent study, Hermans, Aivaloglou,
	and Jansen probed the Enron dataset for uses of VLOOKUP and HLOOKUP
	specifically. The issue at hand was the optional fourth argument, which
	requires that rows are sorted lest it return inaccurate results
	\cite{hermans2015detecting}. For someone seeking to increase awareness of this
	problem, it first allows them to answer the question of significance -- "How
	often do spreadsheet programmers use this function, with these particular
	parameters?" Though \toolname does not discernd whether neighboring columns are
	sorted, it can nevertheless guide the educator to several examples in
	production where this problem surfaces. \par

	\begin{figure}[h] \centering \includegraphics[width=.5\textwidth]{vlookup}
		\includegraphics[width=.5\textwidth]{hlookup} \label{fig:vhlookups}
		\caption{VLOOKUP and HLOOKUP with four arguments.} \end{figure}

	\section{User Study} \label{sec:userstudy} During development, we conducted a
	brief, four-participant user study in order to detect conflicts between our
	design and the goals in section \ref{goals}. We designed the study to be
	exploratory; given trees based in the Enron dataset and at least twenty
	minutes, we asked the participants to adopt the persona of a consultant
	evaluating a company's spreadsheet practice, allowing for questions like which
	functions on which the employees most depended or which sheets or employees
	presented the most anomalous or dangerous designs. Though we instructed them on
	how to use the tool and suggested that they visit one of the most populous
	trees in the set, we put aside any specific tasks or questions and let them
	choose their own paths. Though interpretation of the phrase ``spreadsheet
	practice" can certainly vary from participant to participant, we readily
	accepted this interpretive flexibility since we wanted a range of views and
	backgrounds despite the persona. We clarified, furthermore, that their
	observations could be directed either at the quality of the data conveyed by
	the tool or at the tool's design itself. \par

@@

	\subsection{Informative Qualification} As we found in the case study, the tool
	excels at making formula examples available throughout a tree. Though the tool
	cannot outright explain what a formula does function-by-function, users
	nevertheless found it valuable for the way it offers them with numerous
	practical examples of functions previously unknown or confusing to them,
	especially within the LOOKUP family. In the user study (Section
	\ref{sec:userstudy}), furthermore, we also observed how a participant used the
	repeated instance of a spreadsheet author's name to identify a potential
	problem with a particular designer rather than a design itself, demonstrating
	its use on several levels. \par

	However, the benefits described in the previous sections imply a hidden
	complication: if the tool educates through example but does not explicitly
	distinguish between good design and bad design, how can we be sure that people
	won’t inadvertently learn bad design from it? In the user study, we discussed
	its use in education and smell detection by bringing up the issues surrounding
	the LOOKUP functions; this discussion, though, presupposes a distinguishing eye
	that knew which examples practiced good design and which exhibited the common
	errors. But a beginner might not know the difference. Robillard and Deline, in
	their study, remarked that good API examples should demonstrate best practices,
	but, right now, we have no way to assure this with \toolname. As with the
	problems in achieving unbiased exploration, correcting for this danger may
	entail a reevaluation of design goals.

	\section{Future Work} For these user and case studies, we relied solely on
	Enron's spreadsheets. However, Enron's focus on finances and energy represents
	only one context in which people use spreadsheets. Other corpora, like EUSES
	and Fuse, draw from other domains and industries, and, as Jansen found in his
	comparison of spreadsheets Enron and EUSES \cite{jansen2015enron}, different
	companies rely on different functions. Though the use of a single corpora
	nevertheless suffices for evaluation, we may yet be able to find more
	interesting applications and uses for the tool in other types of spreadsheet.
	\par

	Additionally, Excel is only one of many different spreadsheet products
	available, chosen for this project because of its prevalence. But other
	spreadsheets, such as Gnumeric\footnote{http://www.gnumeric.org/} or Google
	Sheets\footnote{https://www.google.com/sheets/about/} represent other
	approaches to spreadsheet systems. Even within identical domains, there's no
	guarantee that practices and functions will hold across these varied offerings.
	This, then, warrants further analysis, in which good visualization might serve
	well. \par

@@

	Though our focus was specifically on function combination, there are several
	other elements of a formula's context that we do no currently consider. For
	example, \toolname does not current capture argument combination; in
	\ref{fig:index}, it is impossible to state how often the function MATCH
	occurred in both the second and third arguments or whether they were more often
	paired with references. Furthermore, it is easy to open a tree and see what is
	used inside of a specific function, but finding information on what a function
	is used inside of takes considerably more work -- we see again in
	\ref{fig:index} that MATCH is used inside INDEX, but we cannot know which
	functions have INDEX inside them without searching every other tree. More
	distant concerns might be exploring which functions often occur in adjacent
	cells in the sheet -- whenever INDEX occurs, which functions are seen most
	often in the cells around it?

	%\begin{wrapfigure}{O}{.25\textwidth} \centering
	%	\includegraphics[width=.25\textwidth]{longIF} \label{fig:longif} \caption{The
	%		many possibilities of IF, argument 3, demonstrate the problem of having too
	%		many options.} \end{wrapfigure}

@@

	The design of the tool still has much room for improvement. Though the
	exploratory philosophy fostered certain goals well, such as the ability to
	discover new problematic smells without knowing beforehand what they are, it
	hindered others. Some users, for example, found the trees of popular functions
	to be too crowded to explore well, or that their intention to use the tool to
	examine the good or the bad design exclusively was not fully supported by a
	tool that claimed to know neither. Many of the problems, however, were not
	essential to this design conflict and will be overcome in future iterations of
	design by refining icons and improving maneuverability. \par

	Either way, the visualization of these large datasets represents an important
	step in improving practice overall. By taking these voluminous oceans of data
	offered in spreadsheet corpora and rendering them comprehensible through
	visualization, we seek to draw attention to valuable information that would otherwise
	go unnoticed.
